Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-09-10 latex/Chapters/Puzzles.tex
update

\Subsection{2x2x2 Rubik's -- search space, solvability}

@@

@@

In the upcoming implementation (\ref{sec:Implementation}) and results (\ref{sec:ResultsSP}, \ref{sec:ResRubiks}) sections, I will be discussing various solvers. I would put them broadly in four different categories, by decreasing level of human knowledge that has been assumed and handcrafted in them. Notice that this human-knowledge-required scale is obviously somewhat arbitrary, but good to keep in mind in later (and in particular results) sections.

\paragraph{Domain-specific solvers}
At 4 on the human-knowledge-required scale, I would put my NaiveSolver (handcrafted to solve the \textbf{SP}) as well as the hkociemba (2x2x2) and kociemba (3x3x3) open source implementations which I have integrated in my KociembaSolver. They make heavy use of specific knowledge about the two puzzles, not only about their mechanics, but also about how to actually (provably) solve them.

>>>>

\Subsection{2x2x2 Rubik's -- search space, solvability}
\label{Theory:222RCSSS}

@@

\Subsection{Rubik's -- multi-stage search}
\label{Theory:RCMSS}
Modern Rubik's domain-specific solvers (such as the Kociemba solver which I have used in this project as a comparison point) make use of a multi-stage search approach. Essentially, they just do what speedcubers do by hand, just much faster and better. The general idea is as follows: let us suppose we have a puzzle with a state space $\mathcal{S}$ of size $N$ (some large number) and one goal state $g$. Now imagine that we have discovered a much smaller subset of configurations $\mathcal{S}_{1} \in \mathcal{S}$ (say of size $n \ll N$) such that $g \in \mathcal{S}_{1}$ and such that it is \textit{easy} to check whether or not a given configuration $p \in \mathcal{S}$ belongs to $\mathcal{S}_{1}$. In addition, suppose we also have discovered a set of algorithms $\mathcal{A}_{1}$ (sequence of actions) under which elements of $\mathcal{S}_{1}$ are stable and which spans $\mathcal{S}_{1}$. Finally, suppose that we have a set of algorithms $\mathcal{A}_{2}$ which allow us to (provably) take any given configuration $p \in \mathcal{S}$ into $\mathcal{S}_{1}$ in a bounded number of steps. A viable strategy to solve any puzzle then becomes:
\begin{titlemize}{Multi-stage search}
\item If $p \in \mathcal{S}_{1}$, set $a_{2}$ to be the identity move in $\mathcal{S}$, and jump to step 3.
\item Apply a search algorithm (e.g. A$^{*}$ - or some variant such as IDA$^{*}$ (\cite{IDAWiki}) - restricting eligible actions to $\mathcal{A}_{2}$, until we find a sequence of actions $a_{2}$ such that $a_{2}(p) \in \mathcal{S}_{1}$
\item Apply our search algorithm to $a_{2}(p)$, restricting eligible actions to $\mathcal{A}_{1}$, until we find a sequence of actions $a_{1}$ such that $a_{1}(a_{2}(p)) = g$.
\item $a_{1} \circ a_{2}$ is our multi-stage solution
\end{titlemize}
Notice that this is exactly how beginners and advanced speedcubers solve the Rubik's cube, except they usually have more than 2 stages as described above. $S_{1}$ might for instance be the set of cubes where both lower layers are solved, the cross on the top layer as well as all corners are in their respective places (but maybe not oriented as expected), and then they search within a well known and small set of algorithms how to get to the goal state. A previous stage might consist in $S_{2}$, the set of configurations where the two layers are solved, but the top layer is random, and then they will search for a sequence of moves to get into $S_{1}$, etc \dots

@@

In the upcoming implementation (\ref{sec:Implementation}) and results (\ref{sec:ResultsSP}, \ref{sec:ResRubiks}) sections, I will be discussing various solvers. I would put them broadly in four different categories, by decreasing level of human knowledge that has been assumed and handcrafted in them. Notice that my human-knowledge-required scale is obviously somewhat arbitrary, but good to keep in mind in later (and in particular results) sections.

\paragraph{Domain-specific solvers}
At 4 on the human-knowledge-required scale, I would put my NaiveSolver (handcrafted to solve the \textbf{SP}) as well as the hkociemba (2x2x2) and kociemba (3x3x3) open source implementations which I have integrated in my KociembaSolver. They make heavy use of specific knowledge about the two puzzles (see \ref{Theory:RCMSS}), not only about their mechanics, but also about how to actually (and provably) solve them.

### 2022-09-10 latex/Chapters/Rubiks.tex
update

\Section{2x2x2}

1000 seq 30 shuffles training each target network update seen 1.1\% of all puzzles
199 of the 200 perfectly shuffled got solved well under 1h (median 83 seconds), only one outlier took about 67 minutes.

>>>>

\Section{2x2x2}
All the following experiments on the 2x2x2 \textbf{RC} have been run using my Solver's \textit{performance\_test} (see \ref{sec:codesolvers}): for each level of difficulty, the test present the solvers with the same 100 random configurations (for fairness and to increase the significance of the results). The levels of difficulty (number of scrambles from goal) used here go from 2 to 20 in step of 2. The tests were also run with perfect shuffling (difficulty = $\infty$ on the graphs), corresponding to taking scrambling to $\infty$ as explained in \ref{Theory:222RCSSS}).

\Subsection{BFS}
BFS is the only one for which I used a step of 1 on the difficuly and stopped at 5 as it was clear it would not go any further. Given that the run time and expanded nodes are clearly going to increase linearly, and give where difficulty of 5 took this solver already, there was no point going any further!

\Subsection{Kociemba, A$^{*}$[Kociemba], A$^{*}$[DL[A$^{*}$[Kociemba]]]}
Next, let me discuss Kociemba (\ref{}) and associated experiments. We obviously expect Kociemba to be by far the fastest solver since it is handcrafted with much specific knowledge about the Rubik's cube.

\Subsection{Deep reinforcement learner}
\label{sec:S33DRL}
Next I have run the DeepReinforcementLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency=1000}) or when the MSE falls below 10 basis points of the maximum target cost-to-go (\textit{update\_target\_network\_threshold=1e-3})
\item Run for a maximum of 11,000 epochs (\textit{nb\_epochs=11000}) or when the maximum target so far has not been increasing by more than 1\% in the last 5,500 epochs (\textit{max\_target\_uptick=1e-2} and \textit{max\_target\_not\_increasing\_epochs\_pct=0.5})
\item At every update of the target network (\textit{training\_data\_every\_epoch=False}), generate 100 sequences of puzzles (\textit{nb\_sequences=100}), each comprised of a 50-scramble puzzle back to goal state, with all intermediary puzzles along the path (\textit{nb\_shuffles=50}).
\item Use a fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers of size 600, 300, 100 (\textit{layers\_description=(600,300,100)}) and use torch.optim.RMSProp as optimiser (\textit{optimiser=rms\_prop}) with an exponential scheduler (\textit{scheduler=exponential\_scheduler}) with gamma 0.999 (\textit{gamma\_scheduler=0.9999}).
\end{itemize}
Below I show some of the DeepReinforcementLearner's quantities tracked over the epochs (see \ref{fig:33SPDeepReinforcementLearning}). I have shown only the following quantities:
\begin{itemize}
\item learning rate, which decreases due to the scheduler, but gets reset at each network update.
\item max target. One interesting dynamic during learning is that due to the way targets are constructed using value-iteration update, and starting from an all 0 network, the max target only increases by about 1 at best at every network update. That is, the first network only learns how to differentiate goal from non-goal, then the second one learns how to differentiate goals, cost 1 and cost 2+ in some sense. This is why I have introduced the first set of parameters discussed just above. The first one makes sure we do not waste time initially training the left-hand-side network when training is easy (as consist of just goal versus not goal). The second ones make sure that we stop training altogether when the max target does not grow anymore. This is because we do not know a-priori what the God number of a given puzzle/dimension is, but let's assume for the sake of the argument that it is 15. Then that means that after 15 updates of the target network we might not really get so much benefit by keeping training and value-iterating.
\item the MSE loss
\item the MSE loss divided by the max target. The exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 0.01 means little if the we do not compare it to the possible range that the cost-to-go can take.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the DeepReinforcementLearner. This is an interesting quantity, as it tells us later on when testing out of sample and comparing with other solvers, how well the solvers are able to generalise. In this case (3x3) we can see that the DeepReinforcementLearner quickly has seen a very large percentage of all puzzles. That being said, it is good to keep in mind that \textit{seen} does not mean really much here, since the learning is unsupervised. This is quite in contrast to the DeepLearner, for which puzzles seen come along with the cost-to-go computed by a \textit{teacher}, usually an optimal or at least efficient solver.
\end{itemize}

\begin{figure}[H]
\centering
\includegraphics[align=c, scale=0.42]{./Figures/33SPDeepReinforcementLearning}
\caption[33SPDeepReinforcementLearning]{\textbf{DRL} of the 3x3 \textbf{SP}}
\label{fig:33SPDeepReinforcementLearning}
\end{figure}

Notice that the DeepLearner and DeepQLearner have very similar sets of parameters and track similar metrics during their learning for later display, but I shall not detail it here since there is not much extra insight to be had compared to the DeepReinforcementLearning example we just went through.

\Subsection{Solvers' comparison}
